Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-14 paper/ms.tex
yet more refs

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars (CITE), active galactic nuclei (CITE),
and X-ray binaries (CITE).
They have also been used as models for the cosmic microwave background (CITE),
correlated instrumental noise (CITE), spectroscopic calibration (CITE) and
residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~---~this cost is prohibitive.

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly:2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011}, and modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}.

@@

There are three primary applications we envision for the \celerite\ formalism: 1).\ modeling
the variability of an astrophysical system to infer it properties; 2).\
accounting for astrophysical variability as a source of noise when trying to detect
additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic variation of a star
\citep{Corsaro:2014}, measurement of granulation noise \citep{Bastien:2013,Kallinger:2016},
detection of quasi-periodic variability in a high-energy source \citep{McAllister:2016},
or classification of variable objects \citep{Zinn:2016}.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements \citep{Haywood:2014,Rajpaul:2015}. Examples of the third are 
measuring time delays in
multiply-imaged gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of AGN.

>>>>

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic nuclei \citep{Kelly:2014},
and X-ray binaries (CITE).
They have also been used as models for the cosmic microwave background (CITE),
correlated instrumental noise (CITE), spectroscopic calibration (CITE) and
residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~---~this cost is prohibitive.

@@

Although we have in mind application of this fast method to stellar variability,
the method is general for one-dimensional GP problems, and may be applied to
other problems.  Accreting black holes show time series which
may be modeled by correlated noise \citep{Kelly:2014};  indeed, this was the
motivation for the original technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011,Pancoast:2014}, and modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}.

@@

There are three primary applications we envision for the \celerite\ formalism: 1).\ modeling
the variability of an astrophysical system to infer it properties; 2).\
accounting for astrophysical variability as a source of noise when trying to detect
additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic variation of a star
\citep{Brewer:2009,Corsaro:2014}, measurement of granulation noise \citep{Bastien:2013,Kallinger:2016},
detection of quasi-periodic variability in a high-energy source \citep{McAllister:2016},
or classification of variable objects \citep{Zinn:2016}.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements \citep{Haywood:2014,Rajpaul:2015}. Examples of the third are 
measuring time delays in
multiply-imaged gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of AGN.
